---
title: "Race Analysis"
output:
  html_document:
    theme: lumen
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: paged
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })
---

## Setup 

```{r setup, echo=T, results='hide', message=F, warning=F}
library(tidyverse)
library(sf) # fix annoying tidyverse tibble issue with sfc_MULTIPOLYGON
library(clipr)
```


## Import data

import tables from our cleaned data. We'll need both our county-only table and all-terminations

### County Data (with geometry)

```{r import_county_data}
county_terminations_2020 <- read_rds("data-processed/01-county-terminations-2020.rds")

county_terminations_2020 %>% arrange(total_terminations %>% desc())
```

### All terminations

Including out of state and unknown counties

```{r import_all_terminations}
all_terminations <- read_rds("data-processed/02-all-terminations.rds") ## table including out of state and country rows

all_terminations %>% arrange(year)
```

### Totals reported

```{r import_totals}
total_terminations <- read_rds("data-processed/03-itop-totals.rds")

total_terminations
```

## Goals

- Assess counties with highest termination rates on (may need to do this on average) √
- What groups have the highest termination rates? √
- How do those compare with those group's share of the population? √
- Were there any noticeable trends in Texas abortion rates before the restriction was passed this year? (e.g. were they dropping, staying the same, increasing etc) √
- What races were most likely to leave the state for procedures? Who were the least likely? √
- What percentage of abortions occurred outside the state vs inside in 2020? What about 2019 (because COVID may have mattered here) √

## Highest rates in 2020

What counties had the highest rates in 2020?

```{r highest_rates}
terminations_rates_2020 <- county_terminations_2020 %>% 
  arrange(
    terminations_per_pop %>% desc()
  )

terminations_rates_2020
```

## Plot 2020 rates

```{r map_2020_rates}
terminations_rates_2020_plot <- terminations_rates_2020 %>% 
  ggplot() +
  geom_sf(aes(fill = terminations_per_pop), color = "white", size = .2) +
  theme_void() +
  labs(
    title = "Terminations per 1,000 in Texas counties",
    caption = "Source: Texas Health and Human Services"
  ) +
  scale_fill_distiller(
    palette = "Greens",
    direction = 1,
    name = "Cases per 1000"
  ) +
  theme(plot.margin = unit(c(10,0,10,0), "pt"))

terminations_rates_2020_plot
```

## Export for datawrapper

### Step one

Select only the data we need so we can clean it up and export for datawrapper

```{r datawrapper_export}
datawrapper_rates <-  terminations_rates_2020 %>% 
rename(
    county = name
  )  %>% 
  mutate(
    county = str_to_title(county)
  ) %>% 
  select(
    county,
    geoid,
    terminations_per_pop
  ) %>% 
  arrange(county)

datawrapper_rates %>% write_csv("data-processed/datawrapper.csv") # export to csv so we can hackily get rid of geometry in the next step
```

### Step 2

```{r reimport_datawrapper}
data_wrapper_clean <- read_csv("data-processed/datawrapper.csv") %>% 
  select(-geometry) # get rid of geometry

data_wrapper_clean %>% write_clip(allow_non_interactive = T)
```

### Datawrapper chart for story:

Link to datawrapper chart here: [2020 abortion rates](https://datawrapper.dwcdn.net/Hnv4C/1/)

**Question**: 
what's going on in Loving, King, and Terrell counties? 

**Answer**:
Loving has a tiny population (64 people), and in 2020 there were 2 abortions performed there. King, with a population of 265 had 1 abortion, and in Terrell (pop 760), there were 3 abortions in 2020.

**follow-up question**: What were those county's numbers the other years? Let's take a look!

```{r loving_county}
county_terminations_2020 %>% 
  filter(
    name == "LOVING" | name == "TERRELL" | name == "KING"
  ) %>% 
  group_by(name)
```

### Loving and Terrell populations

Let's look at how these counties rank in terms of size compared to other counties

```{r texas_populations}
county_terminations_2020 %>% 
  arrange(total_pop)
```


**Takeways:** 

Loving's 2020 rate was definitely a fluke, but Terrell has had relatively high rates of 8 (2017), 5 (2018), and 4 (2020) which is kind of interesting given there isn't an abortion clinic there. King had a rate of 4 for both 2020 and 2019, but rates of zero for the three years prior.

Interestingly, **all three counties are in the 10 least-populace counties of Texas**: Loving ranking #1 least-populace, King #2 least-populous, and Terrell #7. 


## Abortions over time

How have the number of abortions changed over the ten years, if at all? Analyze then copy for datawrapper

```{r terminations_over_time}
terminations_over_time <- all_terminations %>% 
  group_by(year) %>% 
  summarise(total_terminations = sum(total_terminations)) # get total terminations by year

terminations_over_time %>% write_clip(allow_non_interactive = T) # for datawrapper

terminations_over_time
```

### Plot terminations over time

```{r time_bar_chart}
ggplot(terminations_over_time, aes(x = year, y = total_terminations)) + # set x and y axis
  geom_col(aes(fill = year)) + # add columns
  labs(
    title = "Abortions Trends over the last Decade",
    subtitle = "Before the 6 week ban, abortions in Texas had decline nearly 30% over the last decade",
    caption = "By Ana Goodwin",
    x = "Year",
    y = "Total Terminations"
  )
```


Texas abortions have stayed pretty consistent the last 5 years, only fluctuating by around 3,400. However, abortions have steadily been dropping over time. They have dropped by 28% since 2010.


## Race and terminations

Analyze the following:

- Were any specific races over-represented in terminations in 2020?
- Have any groups seen major increases or decreases in their numbers the last decade?

```{r terminations_by_race}
terminations_by_race <- all_terminations %>% 
  select(
    -(total_terminations:age_not_stated) #get rid of age columns
  ) %>% 
  pivot_longer(
    cols = asian:race_not_stated, # select columns to pivot
    names_to = "race", # new column name
    values_to = "terminations"
  ) %>% 
  mutate(
    race = recode(
      race,
      "other" = "other/unknown",
      "race_not_stated" = "other/unknown",
      "native_american" = "native american"
    )
  ) %>% 
  group_by(year, race) %>% # group by race
  summarise(
    total_terminations = sum(terminations) # add all the termination counts together
  )

terminations_by_race
```

## Plot race data

```{r plot_races}
ggplot(terminations_by_race, aes(x = year, y = total_terminations)) + # set x and y axis
  geom_line(aes(group = race,  color = as.character(race))) + # split lines and set colors
  labs(
    title = "Texas Abortions By Race",
    subtitle = "Reported abortions performed on Texas residents for the years 2016-2020",
    x = "Year",
    y = "Total terminations",
    color = "Year", # fix the title for the legend
    caption = "Source: Texas Health and Human Services"
  )
```

### Race breakdown bar chart

The race data would be better represented by a bar chart

```{r race_bar_chart}
ggplot(terminations_by_race, aes(x = year, y = total_terminations, fill = race)) + 
  geom_bar(stat = "identity") +
  guides(fill = guide_legend(title = "Race/Ethnicity"))
```

### Export this data for datawrapper

To make a grouped bar chart for datawrapper, we need to have this be in a slightly different form

```{r terminations_by_race_datawrapper}
terminations_by_race_datawrapper <- terminations_by_race %>% 
  pivot_wider(
     names_from = race, values_from = total_terminations
  )
  
terminations_by_race_datawrapper %>% write_clip(allow_non_interactive = T)
```

### Datawrapper chart for story:

Link to datawrapper chart here: [abortions by race](https://datawrapper.dwcdn.net/E9f9A/1/)

### Takeaways:

- In 2020, Black women accounted for 30% (16942 terminations) of Texas' total reported abortions for the year (56358 total) 
- In 2020, Hispanic women accounted for 36% (20348 terminations) of Texas' total reported abortions for the year (56358 total) 
- In 2020, White women accounted for (14864	terminations) of Texas' total reported abortions for the year (56358 total) 

### Follow-up questions:

- What percentage of women in Texas are Black and Hispanic?
- How will these groups be impacted by needing to leave the state for procedures? (for interviews)
- What percentage do these groups makeup of out of state abortions for 2020?

## Demographic data

Below are some data points pulled from the American Community Survey 5-year estimates from [the U.S. Census Bureau](https://data.census.gov/cedsci/table?q=Table%20B01001&t=Age%20and%20Sex&g=0400000US48&y=2019&tid=ACSDT5Y2019.B01001&hidePreview=true) for some additional context. 

Texas female population demographics:

- Total female population in Texas: 14,226,847 [data here](https://censusreporter.org/data/table/?table=B01001&geo_ids=04000US48&primary_geo_id=04000US48#valueType|estimate)
- Hispanic females: 5,521,513 [See data here ](https://data.census.gov/cedsci/table?q=Table%20B01001I&t=Age%20and%20Sex&g=0400000US48&y=2019&tid=ACSDT5Y2019.B01001I&hidePreview=true)
- Black females: 1,771,632 [See data here](https://data.census.gov/cedsci/table?q=Table%20B01001b&t=Age%20and%20Sex&g=0400000US48&y=2019&tid=ACSDT5Y2019.B01001B&hidePreview=true)
- White females: 10,502,123 [See data here](https://data.census.gov/cedsci/table?q=Table%20B01001a&t=Age%20and%20Sex&g=0400000US48&y=2019&tid=ACSDT5Y2019.B01001A&hidePreview=true)

## Out of state abortions

First let's get the total number of reported abortions on Texans out of state

```{r other_state_totals}
all_terminations %>% 
  filter(
    county == "OTHER STATE"
  )  %>% 
  group_by(year) %>% 
  summarise(
    OOS_terminations = sum(total_terminations) # add all the termination counts together for Out of State (OOS) totals
  )
```

*takeaway*: for 2020, there were 1124 reported terminations performed on Texas residents. In 2019, less women had abortions performed in different states with 1074.

## OOS terminations by race

Let's see what groups what share of out of state terminations each group had

```{r out_of_state_terminations}
all_terminations %>% 
  filter(
    county == "OTHER STATE"
  ) %>% 
  select(
    -(total_terminations:age_not_stated) #get rid of age columns
  ) %>% 
  pivot_longer(
    cols = asian:race_not_stated, # select columns to pivot
    names_to = "race", # new column name
    values_to = "terminations"
  ) %>% 
  mutate(
    race = recode(
      race,
      "other" = "other/unknown",
      "race_not_stated" = "other/unknown",
      "native_american" = "native american"
    )
  ) %>% 
  group_by(year, race) %>% # group by race
  summarise(
    total_terminations = sum(terminations) # add all the termination counts together
  )

```

Follow up question for interviews: how does TX HHS handle reporting from other states? (AKA how much can we trust this data and how complete is it?)

Takeaways:
- Total terminations performed in different states: 1124
- Hispanic OOS abortions: 120	(10.6%)
- Black OOS abortions: 543 (48%)
- White OOS abortions: 402 (35.6%)

## Facts for story:

1. Hispanic women make up about 39%  of the female population in Texas, according to data from the U.S. Census Bureau, and make up a proportionate share of the terminations for 2020 at 36%. 
2. On the other hand, Black women only make up for about 12.5% of the female population of Texas, but accounted for 30% of the terminations in 2020. 
3. Though Hispanic women made up the largest share of Texas's reported abortions in 2020, they made up only 10% of the abortions performed in other states.
4. Texas abortion numbers have stayed relatively consistent the last 5 years with the highest total being reported in 2019 with 57929 terminations, and the lowest in 2017 with 53843
5. The least populace county in Texas, Loving, had the highest abortion rate per thousand people in 2020. Loving County has a population of only 64 people, but had two abortions in 2020. The second least-populous county, King, had the second-highest rate -- 4 per 1000 -- in 2020 with 1 abortion for its 265 residents. 
6. Five of the least populace counties-- Loving, King, Kenedy, McMullen and Terrell -- were in the top-10 highest abortions rates for the state in 2020. Others in the top-10 were less surprising -- Harris, Bexar, Dallas, and Travis had some of the highest rates but are also areas with some of the largest populations in the state and more access to abortion providers.  
7. White women accounted for 26% of abortions in the state, despite representing 74% of the female population.
8. Abortions decreased by 28% between 2010 and 2020

## Up Next: [Age Analysis](03-age-analysis.html)

